---
title: "CDC Data Exercise"
---

This dataset of U.S. mortality trends since 1900 highlights the differences in age-adjusted death rates and life expectancy at birth by race and sex. The dataset contains 1,071 observations along with five columns. The five variables are `year`, `race`, `sex`, `average life expectancy`, and `age-adjusted death rate`.

The dataset can be downloaded [here](https://data.cdc.gov/NCHS/NCHS-Death-rates-and-life-expectancy-at-birth/w9j2-ggv5/about_data "Death rates and life expectancy at birth").

```{r, warning=FALSE, message=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
```

```{r}
#load the file
death_rate <- read_csv("NCHS_-_Death_rates_and_life_expectancy_at_birth_20240702.csv")
```

```{r}
head(death_rate)
```

```{r}
#Remove "Both Sexes" from the "Sex" column and keep only male and female
death2 <- death_rate %>%
  filter(Sex != "Both Sexes")
```

```{r}
#View the new dataset to ensure correct variables
View(death2)
```

```{r, warning=FALSE}
# Summary statistics for Average Life Expectancy
life_expectancy_summary <- summary(death2$`Average Life Expectancy (Years)`)
print(life_expectancy_summary)

# Histogram of Average Life Expectancy
AvgLifeExp_plot <- ggplot(death2, aes(x = `Average Life Expectancy (Years)`)) +
  geom_histogram(binwidth = 1, fill = "orange", color = "black") +
  labs(title = "Distribution of Average Life Expectancy", x = "Average Life Expectancy (years)", y = "Count")
AvgLifeExp_plot
```

```{r}
# Summary statistics for Age-adjusted Death Rate
death_rate_summary <- summary(death2$`Age-adjusted Death Rate`)
print(death_rate_summary)

# Histogram of Age-adjusted Death Rate
ggplot(death2, aes(x = `Age-adjusted Death Rate`)) +
  geom_histogram(binwidth = 50, fill = "red", color = "black") +
  labs(title = "Distribution of Age-adjusted Death Rate", x = "Age-adjusted Death Rate", y = "Count")

```

```{r, warning=FALSE}
#plot with gradient color scale
AvgLife_Race_plot <- ggplot(death2, aes(x = Race, y = `Average Life Expectancy (Years)`, color = `Average Life Expectancy (Years)`)) +
  geom_point(size = 5) +
  scale_color_gradient(low = "red", high = "green") +
  labs(title = "Average Life Expectancy by Race", x = "Race", y = "Average Life Expectancy (Years)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
AvgLife_Race_plot
```

